---
title: "Colony counts"
author: "Julianna Renzi"
date: "1/25/2023"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


require(here)
require(tidyverse)
require(vegan)
require(lme4)
```

# Read in data

```{r}
# colony counts
colonies <- read_csv(here("data/Colony_counts.csv")) %>% 
  # fix this weird date entry
  mutate(Date = case_when(
    Date == "8/21/22" ~ "8/21/2022",
    TRUE ~ Date
  ))

# bring in site visual nutrient file
nuts <- read_csv(here("data/site_metadata.csv")) 
  
# and nutrient working group file that we will use as metadata
nutMetadat <- read_csv(here("data/All_dat_withFish.csv")) 
# Make overall site stats
```

What are the misspelt coral genera?

```{r}
colonies %>% 
  select(Genus) %>% 
  unique() 

# fix them
colonies %>% 
  mutate(Genus = case_when(
    Genus == "Cyphestrea" ~ "Cyphastrea",
    Genus == "Fungiid" ~ "Fungiidae",
    Genus == "leptastrea" ~ "Leptastrea",
      Genus == "Leptastrea?" ~ "Leptastrea",
      Genus == "Napropora" ~ "Napopora",
      Genus == "Napropora" ~ "Napopora",
    Genus == "Napapora"~ "Napopora",
      Genus == "Pammacora" ~ "Psammocora",
      Genus == "Psammacora" ~ "Psammocora", 
      Genus == "Porites rus" ~ "Porites_rus",
      Genus == "Porites_Rus" ~ "Porites_rus",
      Genus == "Porites_Massive" ~ "Porites_massive",
    TRUE ~ Genus
  )) %>% 
  filter(Genus != "Single_polyp") -> colonies_cleaned


# check
colonies_cleaned %>% 
  select(Genus) %>% 
  unique() 
```


# Make long df

```{r}
colonies_cleaned %>% 
  mutate(Transect = parse_number(Transect)) %>% 
  mutate(TransectSite = paste0(Site, "T", Transect)) %>% 
  group_by(Transect, TransectSite, Date, Site, Genus) %>% 
  summarize(Abundance = sum(Count)) %>% 
  ungroup() %>% 
  # replace this with the nearest station
  mutate(Site = ifelse(Site == "Gump", 159, Site)) %>% 
  mutate(Site = as.double(Site)) %>% 
  # join visual nutrient data
  full_join(nuts, by = "Site") -> genusAbund_temp

```

Then want to join it with key parts of the nutrient working group metadata:

```{r}
# select the stuff we want:
nutMetadat %>% 
  filter(TimePeriod == "2021-May") %>% 
  mutate(Site = as.double(Site)) %>% 
  select(Site, Latitude, Longitude, Distance_to_crest, Distance_to_shore, Distance_to_pass, Distance_to_deep_lagoon_water, Distance_to_population_center, Habitat, Microbial_Species_Richness, Microbial_Shannon_Diversity, Microbial_Phylogenetic_Diversity, Microbial_Evenness, Microbial_PCoA1, Microbial_PCoA2, FishNH4ug_min_m2, FishBiomg_m2) -> nmwgMetadat

# also do a separate one for 124 since it doesn't have a May 2021 line in the df:
nutMetadat %>% 
  filter(Site == 124) %>% 
  filter(TimePeriod == "2020-August") %>% 
  mutate(Site = as.double(Site)) %>% 
  select(Site, Latitude, Longitude, Distance_to_crest, Distance_to_shore, Distance_to_pass, Distance_to_deep_lagoon_water, Distance_to_population_center, Habitat, Microbial_Species_Richness, Microbial_Shannon_Diversity, Microbial_Phylogenetic_Diversity, Microbial_Evenness, Microbial_PCoA1, Microbial_PCoA2, FishNH4ug_min_m2, FishBiomg_m2) -> nmwgMetadat124
```

Join with long df:

```{r}
genusAbund_temp %>% 
  filter(Site != 124) %>% 
  left_join(nmwgMetadat) -> genusAbund_tempNo124
  
genusAbund_temp %>% 
  filter(Site == "124") %>% 
  left_join(nmwgMetadat124, by = "Site") -> genusAbund_temp124
  
  
# put them together
genusAbund_tempNo124 %>% 
  filter(Site != 163) %>% 
  bind_rows(genusAbund_temp124) -> genusAbund_Long
```

# Look at individual general and total abundance

Get all ID df

```{r}
genusAbund_Long %>% 
  group_by(TransectSite, Date, Transect, Site) %>% 
  summarize(N = n()) %>% 
  select(-N) -> all_ids
```


## Montipora

Make Montipora df

```{r}
genusAbund_Long %>% 
  filter(Genus == "Montipora") %>% 
  full_join(all_ids) %>% 
  mutate(Abundance = replace_na(Abundance, 0)) -> montipAbund

```


Plot

```{r}
montipAbund %>% 
  ggplot(aes(x = Nutrient_estimate, y = Abundance, color = as.factor(Site))) +
  geom_boxplot() +
  geom_point() +
  theme_classic()
```


```{r}
montipAbund %>% 
  ggplot(aes(x = FishBiomg_m2, y = Abundance, color = as.factor(Site))) +
  geom_boxplot() +
  geom_point() +
  theme_classic()
```

```{r}
m.Montip <- glmer(Abundance ~ Nutrient_estimate + scale(FishBiomg_m2) + (1 | Site), data = montipAbund, family = poisson(link = "log"))
  summary(m.Montip)
  plot(m.Montip)
```

### Size

```{r}
colonies_cleaned %>% 
  mutate(Transect = parse_number(Transect)) %>% 
  mutate(TransectSite = paste0(Site, "T", Transect)) %>% 
  group_by(Transect, TransectSite, Date, Site, Genus, Size_cm) %>% 
  summarize(Abundance = sum(Count)) %>% 
  ungroup() %>% 
  # replace this with the nearest station
  mutate(Site = ifelse(Site == "Gump", 159, Site)) %>% 
  mutate(Site = as.double(Site)) %>% 
  # join visual nutrient data
  full_join(nuts, by = "Site") %>% 
  filter(Genus == "Montipora") %>% 
  full_join(all_ids) %>% 
  mutate(Abundance = replace_na(Abundance, 0)) %>% 
  mutate(Size_cm = factor(Size_cm, levels = c("<5", "5-10", "10-20", "20-40", "40-80", "80-160", ">160"))) -> montipSize
```

```{r}
montipSize %>% 
  #group_by(Size_cm, Nutrient_estimate) %>% 
  #summarize(Mean_abund = mean(Abundance),
  #          sd_abund = sd(Abundance)) %>% 
  ggplot(aes(x = Size_cm, y = Abundance)) + 
  geom_boxplot() +
  geom_point(aes(color = as.factor(Site))) +
  facet_wrap(~as.factor(Nutrient_estimate)) +
  theme_bw()
```


## Pocillopora

Make Pocillopora df

```{r}
genusAbund_Long %>% 
  filter(Genus == "Pocillopora") %>% 
  full_join(all_ids) %>% 
  mutate(Abundance = replace_na(Abundance, 0)) %>% 
  rename(Poc_abund = Abundance )-> pocAbund

```


Plot

```{r}
pocAbund %>% 
  ggplot(aes(x = Nutrient_estimate, y = Poc_abund, color = as.factor(Site))) +
  geom_boxplot() +
  geom_point() +
  theme_classic()
```


```{r}
pocAbund %>% 
  ggplot(aes(x = FishBiomg_m2, y = Poc_abund, color = as.factor(Site))) +
  geom_boxplot() +
  geom_point() +
  theme_classic() # Huh... Wonder if that's real? 
```

```{r}
m.Poc <- glmer(Poc_abund ~ Nutrient_estimate + scale(FishBiomg_m2) + (1 | Site), data = pocAbund, family = poisson(link = "log"))
  summary(m.Poc)
  plot(m.Poc)
```

### Size

```{r}
colonies_cleaned %>% 
  mutate(Transect = parse_number(Transect)) %>% 
  mutate(TransectSite = paste0(Site, "T", Transect)) %>% 
  group_by(Transect, TransectSite, Date, Site, Genus, Size_cm) %>% 
  summarize(Abundance = sum(Count)) %>% 
  ungroup() %>% 
  # replace this with the nearest station
  mutate(Site = ifelse(Site == "Gump", 159, Site)) %>% 
  mutate(Site = as.double(Site)) %>% 
  # join visual nutrient data
  full_join(nuts, by = "Site") %>% 
  filter(Genus == "Pocillopora") %>% 
  full_join(all_ids) %>% 
  mutate(Abundance = replace_na(Abundance, 0)) %>% 
  mutate(Size_cm = factor(Size_cm, levels = c("<5", "5-10", "10-20", "20-40", "40-80", "80-160", ">160"))) -> pocSize
```

```{r}
pocSize %>% 
  #group_by(Size_cm, Nutrient_estimate) %>% 
  #summarize(Mean_abund = mean(Abundance),
  #          sd_abund = sd(Abundance)) %>% 
  ggplot(aes(x = Size_cm, y = Abundance)) + 
  geom_boxplot() +
  geom_point(aes(color = as.factor(Site))) +
  facet_wrap(~as.factor(Nutrient_estimate)) +
  theme_bw()
```


### Poc v. Montip?

```{r}
montipAbund %>% 
  rename(Montip_abundance = Abundance) %>% 
  select(-Genus) %>% 
  full_join(pocAbund) -> WRONG
```

AHH THIS IS FUCKED UP I GOTTA FIX THIS. Doing this next bit because I want closure before I move on

```{r}
WRONG %>% 
  ggplot(aes(x = Montip_abundance, y = Poc_abund, color = (Nutrient_estimate))) +
  geom_point() +
  # 1:1 line
  geom_abline(slope = 1, intercept = 0) +
  scale_colour_gradient(low = "gray", high = "red") +
  theme_classic()
```

```{r}
mPocMonti <- glm(Poc_abund ~ Montip_abundance + Nutrient_estimate + Montip_abundance:Nutrient_estimate, family = poisson, data = WRONG)
  summary(mPocMonti)
  plot(mPocMonti)
```

Hmmm... What do the predictions say?

```{r}
pred <- expand.grid(Montip_abundance = seq(from = min(WRONG$Montip_abundance, na.rm = TRUE),
                         to = max(WRONG$Montip_abundance, na.rm = TRUE),
                         length.out = length(WRONG$Montip_abundance)),
                    Nutrient_estimate = c(min(WRONG$Nutrient_estimate, na.rm = TRUE),
                                          mean(WRONG$Nutrient_estimate, na.rm = TRUE),
                         max(WRONG$Nutrient_estimate, na.rm = TRUE)))



pred$PredPocAbund <- predict(mPocMonti, 
                 pred, 
                 type = "response")


# plot
pred %>% 
  ggplot(aes(x = Montip_abundance, y = PredPocAbund, color = Nutrient_estimate)) +
  #ylab("Predicted Pocillopora abundance") +
  scale_colour_gradient(low = "gray", high = "red") +
  geom_point() +
  theme_classic()


  # ggsave(here("figures/PredTrapeziidabund.jpg"), width = 4.5, height = 3, dpi = 500)
```


What about little Pocillopora recruits vs. Montipora cover??????
Also is this done right? Do we need zeros for each size class??????



## Get spp matrix

```{r}
genusAbund_Long %>% 
  pivot_wider(names_from = Genus, values_from = Abundance) %>% 
  # replace NA's with zeros
  mutate_all(~replace(., is.na(.), 0)) -> spp_matrix_wide # for no coral_ID remove beginning columns
```


Deal with rare genera

```{r}
spp_mat_rare <- spp_matrix_wide[, colSums(spp_matrix_wide != 0) >= round(dim(spp_matrix_wide)[1]*0.05)] # at least 5% of samples
```

# Look at diversity metrics


```{r}
# get index for where coral columns start
first_coral <- grep("Acropora", colnames(spp_mat_rare))

shann.in <- diversity(spp_mat_rare[,first_coral:dim(spp_mat_rare)[2]], index = "shannon")
simp.in <- diversity(spp_mat_rare[,first_coral:dim(spp_mat_rare)[2]], index = "simpson")
invSimp.in <- diversity(spp_mat_rare[,first_coral:dim(spp_mat_rare)[2]], index = "invsimpson")
spp.rich <- specnumber(spp_mat_rare[,first_coral:dim(spp_mat_rare)[2]])



div.metrics <- cbind(shann.in, simp.in, invSimp.in, spp.rich, spp_mat_rare[,1:(first_coral-1)]) 
  

```

## By habitat

```{r}
# shannon
div.metrics %>% 
  ggplot(aes(x = as.factor(Site), y = shann.in, fill = KellyDeron_habitat)) +
  geom_boxplot() +
  ylab("Shannon diversity") +
  theme_classic()

# spp richness
div.metrics %>% 
  ggplot(aes(x = as.factor(Site), y = spp.rich, fill = KellyDeron_habitat)) +
  geom_boxplot() +
  ylab("Species richness") +
  theme_classic()


# simpson
div.metrics %>% 
  ggplot(aes(x = as.factor(Site), y = simp.in, fill = KellyDeron_habitat)) +
  geom_boxplot() +
  ylab("Simpson's index") +
  theme_classic()

# inverse simpson
div.metrics %>% 
  ggplot(aes(x = as.factor(Site), y = invSimp.in, fill = KellyDeron_habitat)) +
  geom_boxplot() +
  theme_classic()

```


## By nutrient regime

```{r}
# shannon
div.metrics %>% 
 #filter(Site != 159) %>% 
  ggplot(aes(x = Nutrient_estimate, y = shann.in, color = as.factor(Site))) +
  geom_boxplot() +
  geom_point() +
  ylab("Shannon diversity") +
  theme_classic() # Gump is just crazy..

# spp richness
div.metrics %>% 
  #filter(Site != 159) %>% 
  ggplot(aes(x = Nutrient_estimate, y = spp.rich, color = as.factor(Site))) +
  geom_boxplot() +
  geom_point() +
  ylab("Coral species richness") +
  xlab("Visual Nutrient Estimate") +
  theme_classic()

ggsave(here("figures/SppRichNut.jpg"), width = 7, height = 4.6, dpi = 500)


# simpson
div.metrics %>% 
  #filter(Site != 159) %>% 
  ggplot(aes(x = Nutrient_estimate, y = simp.in, color = as.factor(Site))) +
  geom_boxplot() +
  geom_point() +
  ylab("Simpson's index") +
  theme_classic()


# inverse simpson
div.metrics %>% 
  #filter(Site != 159) %>% 
  ggplot(aes(x = Nutrient_estimate, y = invSimp.in, color = as.factor(Site))) +
  geom_boxplot() +
  geom_point() +
  theme_classic()

```

## Weird model?

```{r}
m.sppRich <- glm(spp.rich ~ Nutrient_estimate, family = poisson, data = div.metrics)
  summary(m.sppRich)
  plot(m.sppRich)
  
# What about without Gump?
m.sppRichnoGump <- glm(spp.rich ~ Nutrient_estimate, family = poisson, data = subset(div.metrics, Site != 159))
  summary(m.sppRichnoGump) # nope
  plot(m.sppRichnoGump)  

```


## By distance to different things

```{r}
# distance to shore
div.metrics %>% 
  filter(Site != 159) %>% 
  ggplot(aes(x = Distance_to_shore, y = spp.rich, color = as.factor(Site))) +
  geom_boxplot() +
  geom_point() +
  ylab("Species richness") +
  theme_classic()

# distance to crest--kinda interesting
div.metrics %>% 
  filter(Site != 159) %>% 
  ggplot(aes(x = Distance_to_crest, y = spp.rich, color = as.factor(Site))) +
  geom_boxplot() +
  geom_point() +
  ylab("Species richness") +
  theme_classic()


div.metrics %>% 
  filter(Site != 159) %>% 
  ggplot(aes(x = Distance_to_crest, y = spp.rich, color = as.factor(Site))) +
  geom_boxplot() +
  geom_point() +
  ylab("Species richness") +
  theme_classic()


# Distance_to_pass
div.metrics %>% 
  filter(Site != 159) %>% 
  ggplot(aes(x = Distance_to_crest, y = spp.rich, color = as.factor(Site))) +
  geom_boxplot() +
  geom_point() +
  ylab("Species richness") +
  theme_classic()


# Distance_to_population_center
div.metrics %>% 
  filter(Site != 159) %>% 
  ggplot(aes(x = Distance_to_population_center, y = spp.rich, group = as.factor(Site), fill = Nutrient_estimate)) +
  geom_boxplot() +
  geom_point() +
  ylab("Coral pecies richness") +
  xlab("Distance to population center") +
  scale_fill_gradient(low = "gray", high = "red") +
  theme_classic() # interesting maybe?

ggsave(here("figures/SppRichDistPop.jpg"), width = 7, height = 4.6, dpi = 500)



# Microbial_Species_Richness
div.metrics %>% 
  #filter(Site != 159 ) %>% 
  filter(Site != 124) %>% 
  ggplot(aes(x = Microbial_Species_Richness, y = spp.rich, color = as.factor(Site))) +
  geom_boxplot() +
  geom_point() +
  ylab("Coral species richness") +
  xlab("Microbe species richness") +
  theme_classic() # potentially cool!

ggsave(here("figures/SppRichMicrobe.jpg"), width = 7, height = 4.6, dpi = 500)


# Microbial_Shannon_Diversity
div.metrics %>% 
  #filter(Site != 159 ) %>% 
  filter(Site != 124) %>% 
  ggplot(aes(x = Microbial_Shannon_Diversity, y = spp.rich, color = as.factor(Site))) +
  geom_boxplot() +
  geom_point() +
  ylab("Coral species richness") +
  xlab("Microbe shannon diversity")+
  theme_classic() # potentially cool?

ggsave(here("figures/SppRichMicrobeShann.jpg"), width = 7, height = 4.6, dpi = 500)


# Microbial_Phylogenetic_Diversity
div.metrics %>% 
  filter(Site != 159) %>% 
  filter(Site != 124) %>% 
  ggplot(aes(x = Microbial_Phylogenetic_Diversity, y = spp.rich, color = as.factor(Site))) +
  geom_boxplot() +
  geom_point() +
  ylab("Species richness") +
  theme_classic()

# Microbial_Evenness
div.metrics %>% 
  filter(Site != 159) %>% 
  filter(Site != 124) %>% 
  ggplot(aes(x = Microbial_Evenness, y = spp.rich, color = as.factor(Site))) +
  geom_boxplot() +
  geom_point() +
  ylab("Species richness") +
  theme_classic() # cool


# Microbial_PCoA1
div.metrics %>% 
  #filter(Site != 159) %>% 
  filter(Site != 124) %>% 
  ggplot(aes(x = Microbial_PCoA1, y = spp.rich, color = as.factor(Site))) +
  geom_boxplot() +
  geom_point() +
  ylab("Species richness") +
  theme_classic() 



# FishBiomg_m2
div.metrics %>% 
  filter(Site != 159) %>% 
  filter(Site != 124) %>% 
  ggplot(aes(x = FishBiomg_m2, y = spp.rich, color = as.factor(Site))) +
  geom_boxplot() +
  geom_point() +
  ylab("Species richness") +
  theme_classic() 


```


## Other weird model?

```{r}
m.sppRichDist <- glmer(spp.rich ~ Nutrient_estimate + scale(Distance_to_population_center) + Nutrient_estimate:scale(Distance_to_population_center) + (1 | Site), data = div.metrics, family = poisson(link = "log"))  
  summary(m.sppRichDist)
  
  
# microbes?
m.sppRichMicrobe <- glmer(spp.rich ~ scale(Microbial_Species_Richness) + (1 | Site), data =div.metrics, family = poisson(link = "log"))  
  summary(m.sppRichMicrobe)
  

m.sppRichDistHab <- glmer(spp.rich ~ Nutrient_estimate + KellyDeron_habitat + Nutrient_estimate:KellyDeron_habitat + (1 | Site), data = div.metrics, family = poisson(link = "log"))  
  summary(m.sppRichDistHab)
    
  
```

Hmmm... What do the predictions say?

```{r}
pred <- expand.grid(Distance_to_population_center = seq(from = min(div.metrics$Distance_to_population_center, na.rm = TRUE),
                         to = max(div.metrics$Distance_to_population_center, na.rm = TRUE),
                         length.out = length(div.metrics$Distance_to_population_center)),
                    Nutrient_estimate = c(min(div.metrics$Nutrient_estimate, na.rm = TRUE),
                                          mean(div.metrics$Nutrient_estimate, na.rm = TRUE),
                         max(div.metrics$Nutrient_estimate, na.rm = TRUE)),
                    Site = unique(div.metrics$Site))



pred$PredSppRich <- predict(m.sppRichDist, 
                 pred, 
                 type = "response")


# plot
pred %>% 
  ggplot(aes(x = Distance_to_population_center, y = PredSppRich, color = Nutrient_estimate)) +
  geom_point() +
  ylab("Predicted coral species richness") +
  xlab("Distance to population center") +
  scale_color_gradient(low = "gray", high = "red") +
  facet_wrap(~Nutrient_estimate) +
  theme_classic() # interesting maybe?
  



# ggsave(here("figures/PredTrapeziidabund.jpg"), width = 4.5, height = 3, dpi = 500)
```


For habitat:

Hmmm... What do the predictions say?

```{r}
pred <- expand.grid(Distance_to_population_center = seq(from = min(div.metrics$Distance_to_population_center, na.rm = TRUE),
                         to = max(div.metrics$Distance_to_population_center, na.rm = TRUE),
                         length.out = length(div.metrics$Distance_to_population_center)),
                    Nutrient_estimate = c(min(div.metrics$Nutrient_estimate, na.rm = TRUE),
                                          mean(div.metrics$Nutrient_estimate, na.rm = TRUE),
                         max(div.metrics$Nutrient_estimate, na.rm = TRUE)),
                    Site = unique(div.metrics$Site),
                    KellyDeron_habitat = unique(div.metrics$KellyDeron_habitat))



pred$PredSppRich <- predict(m.sppRichDistHab, 
                 pred, 
                 type = "response")


# plot
pred %>% 
  mutate(KellyDeron_habitat = factor(KellyDeron_habitat, levels = c("Fringing_reef", "Mid_lagoon", "Reef_pass"))) %>% 
  ggplot(aes(x = Nutrient_estimate, y = PredSppRich, color = Nutrient_estimate)) +
  geom_point() +
  geom_line() +
  ylab("Predicted coral species richness") +
  xlab("Visual nutrient estimate") +
  scale_color_gradient(low = "gray", high = "red") +
  facet_wrap(~KellyDeron_habitat) +
  theme_classic() # interesting maybe?
  



# ggsave(here("figures/PredTrapeziidabund.jpg"), width = 4.5, height = 3, dpi = 500)
```




